Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device

### 2023-07-07 manuscript/manuscript.qmd
results almost done. XXgboost plots are ok.

---
title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality metrics
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
---

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality metrics, including sleep period time
(SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO). ~~By
analyzing these additional metrics, we aimed to provide a comprehensive
evaluation of the model's capability in assessing various aspects of
sleep quality, allowing us to determine the overall effectiveness of our
models in accurately estimating different parameters related to sleep
duration and quality.~~

@@

The current study leverages data from the SCREENS
project[@rasmussen2020], a study conducted from October 2018 to March
2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we isolated
data from child participants between the ages of 6 to 10 years within
the SCREENS cohort. Our main sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, utilizes advanced EEG hardware and signal
processing algorithms, employs three self-adhesive, disposable sensors
placed outside the hairline for reliable EEG signal acquisition. The ZM
uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is utilized
for accurate sleep detection, showcasing its aptness for in-home
monitoring[@kaplan2014], while the Z-PLUS effectively differentiates
sleep stages, as evidenced by its alignment with expert evaluations
using PSG data[@wang2015].

Finally, we affirm that the SCREENS study received approval from the
Regional Scientific Committee of Southern Denmark, and all data handling
processes complied with the General Data Protection Regulation (GDPR),
guaranteeing the ethical and secure management of participant
information.

@@

In this study, raw data processing began with a low-pass filtration step
using a 4th order Butterworth filter with a 5 Hz cut-off frequency to
eliminate high-frequency noise. Following filtration, data were
partitioned into overlapping 2-second intervals, each successive
interval sharing a 50% overlap with the previous one similar to methods
described by Skotte et al.[@skotte_detection_2014]. Any non-wear data
was remove using previously described methods[@skovgaard2023] and data
was resampled to 30-second epochs so every sample classified by the
algorithms corresponds to a 30-second epoch scored during the ZM
recordings. Subsequently, we performed a feature extraction process that
yielded a set of 88 features, providing a robust characterization of the
data. Extracted from accelerometer and temperature signals, these
features include temporal elements that use both lag and lead values,
capturing dynamic data trends by incorporating measurements from
preceding and upcoming intervals. Furthermore, inspired by Walch et
al.[@walch2019], we developed sensor-independent features to encapsulate
circadian rhythms. These features offer unique insights not directly
discernible from sensor outputs and are meant to approximate the
changing drive of the circadian clock to sleep over the course of the
night (see @fig-sensor-independent). Furthermore, the feature set was
enriched by including signal characteristics, which encompass vector
magnitude, mean crossing rate, skewness, and kurtosis for each of the x,
y, and z dimensions. These features provide a comprehensive
understanding of the signal's attributes, facilitating a more nuanced
analysis of the underlying patterns. All features are summarized in
table ??? (overvej tabel til supp. material)

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings commonly occurring at the end of each
cycle[@galland_normal_2012]. Examining the raw ZM predictions, we noted
a significant overestimation in the number of awakenings per night for
the children in our study, exceeding what would be expected based on
typical sleep cycle patterns. Consequently, we elected to train and
evaluate our models using not only the raw ZM output, but also versions
that were subjected to 5-minute and 10-minute median filters. This
approach resulted in an anticipated, more age-appropriate count of
awakenings per night, providing a more accurate depiction of children's
sleep patterns (see @fig-zm-median).

@@

In our study, we utilize two distinct modeling strategies to predict
sleep patterns derived from thigh-mounted accelerometer data: a
sequential ensemble and a multiclass bidirectional Long Short-Term
Memory (biLSTM)[@hochreiter1997] neural network. The sequential ensemble
involves using four pairs of models, each pair consisting of the same
algorithm, that predict in-bed and sleep time successively. Conversely,
the biLSTM is a comprehensive model that predicts multiple classes
simultaneously, effectively handling complex temporal patterns. By
applying and comparing both, we aim to evaluate their relative
effectiveness.

@@

The biLSTM is a multiclass classifier capable of predicting the classes
of out-bed-awake, in-bed-awake, and in-bed-asleep. The architecture of
the biLSTM consisted of four layers, each with 128 hidden units. The
choice of four layers and 128 hidden units in our model balanced
complexity and efficiency: it was deep enough to learn intricate
patterns yet feasible to train timely. Additionally, the bidirectional
nature of the LSTM, doubling the hidden units at each time step,
improved data comprehension and avoided overfitting.

something on sequence length and step size

The biLSTM model was chosen for sleep classification using accelerometer
data from a thigh-mounted device due to its advantages in modeling
temporal context, handling a large temporal scope, and enabling temporal
inference over any feature without the need for hand-designed temporal
features.

Previous studies have explored the use of LSTM models for sleep
detection, showing promising results in capturing complex temporal
patterns. Notable works by Sano et al. [@sano2019] and Chen et al.
[@chen2021] have demonstrated the potential of LSTM models in improving
sleep detection based on accelerometer data. However, further research
is necessary to fully comprehend the strengths and limitations of this
approach.

@@

For the models in sequence, we trained four pairs of classification
models, each pair distinguishing between in-bed/out-of-bed and
asleep/awake states. The dataset was randomly split into a training set
(approximately 50% of the subjects) and a testing set (also
approximately 50%), ensuring that samples from the same subject were
never simultaneously used in both sets. Hyper-parameter optimization was
performed using 10-fold Monte Carlo cross-validation, with the F1 score
as the performance metric. The best-performing set of hyperparameters
was then used to fit the models to the full training dataset, maximizing
accuracy by leveraging all available data. Furthermore, the dataset for
predicting sleep was highly imbalanced which can pose challenges when
training machine learning models, as models may favor predicting the
majority class. To address this issue, we employed oversampling
techniques to ensure that each class had a roughly equal number of
training samples. Specifically, we utilized the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002], which generates new
samples by interpolating random samples with their nearest neighbors. In
our study, we implemented SMOTE using the themis R package[@themis],
resampling all classes to achieve a balanced distribution of training
samples.

The biLSTM model was trained using the Adam optimizer, which is
computationally efficient and adapts the learning rate during training.
Cross-entropy loss function was employed for its suitability in
multiclass classification with mutually exclusive classes. The softmax
activation function was selected for the output layer to obtain a
probability distribution over the classes. Data for training the biLSTM
were randomly divided into a training, validation, and test tensors
based on a 50/25/25 split. The model was evaluated using the F1 score on
both the training and validation sets. Early stopping was implemented
with a patience of 3 epochs, meaning training would stop if there was no
improvement in the validation loss for 3 consecutive epochs.

### Model Validation \[evt. bare "statistics"\]

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs. out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs. awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. Furthermore, we
present precision-recall curves and receiver operating characteristic
curves, including the area under the curve, for each model and for each
type of ZM prediction.We considered both the in-bed/out-of-bed and
awake/asleep scoring tasks as binary classification problems,
designating in-bed and asleep as the positive labels and out-of-bed and
awake as the negative labels in accordance with previous
research[@hjorth2012; @kushida2001].

To evaluate our developed models' performance in assessing sleep quality
metrics, we employed Bland-Altman plots and Pearson correlations.
Additionally, we used the Bland-Altman method to assess the agreement
between two measurement techniques. Given the multiple observations per
subject in our dataset, we incorporated a bootstrap procedure to account
for this additional source of variability. We computed the mean
difference (bias) and the limits of agreement as the mean difference
±1.96 times the standard deviation of the differences. The 95%
confidence intervals for the bias and the limits of agreement were
estimated using the bias-corrected and accelerated bootstrap method with
10,000 bootstrap replicates, to ensure our measurements are robust and
account for intra-subject variability. These methods were used to
compare the sleep quality summaries calculated by our models with those
obtained from EEG-based ZM sleep quality summaries. The sleep quality
summaries included the following metrics:

@@

Our models were trained to predict sleep quality metrics using the full
span of accelerometer recordings, beyond the ZM recording duration. This
approach, applied to all models and ZM prediction types, allows for
comprehensive sleep pattern analysis on longitudinal accelerometer data.

@@

![Flowchart of eligible nights included in the study - revise text
boxes!](visuals/flowchart_of_elligible_nights.pdf)

The analysis included children with an average age of 9.4 years (SD =
2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
86% of the total recording duration.

As reported in @tbl-zm_overview the sleep quality metrics derived from
ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. The Sleep Period Time (SPT) were consistent
across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as this
corresponds to the length of the ZM recording. Total Sleep Time (TST)
and Sleep Efficiency (SE) increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency
to Persistent Sleep (LPS) also elevated, suggesting the filter smooths
out brief awakenings at sleep onset, leading to a prolonged time to
persistent sleep. The most significant change was seen in Wake After
Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
10-minute filtered data, respectively. The number of awakenings was also
considerably reduced with the application of filters. In the raw data,
the average number of awakenings was 34.46 ± 11.33 per night, which
reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
filtered data sets respectively. These results underscore the role of
median filters on sleep metrics.

@@

## Performance on Epoch-To-Epoch Basis

The epoch-to-epoch evaluation of in-bed performance metrics, outlined in
@tbl-in_bed_performance, demonstrates practically equivalent performance
across all model types. While the Decision Tree model posted an F1 score
of 94.4% and accuracy of 95.3%, the Logistic Regression and Feed-Forward
Neural Network models each exhibited an F1 score of 95.0% and similar
accuracies of 95.7% and 95.8%, respectively. The XGBoost model, despite
recording the highest metrics with an F1 score of 95.4% and accuracy of
96.1%, outpaced the others only marginally. This underscores the
consistency of performance among these models in classifying in-bed
conditions.

@@

@tbl-sleep_performance details the performance of the sequential models
on raw and median-filtered (5 and 10 minute) ZM predictions. The F1
Scores, which are unweighted macro averages, for raw ZM predictions
range from 71.05% to 76.18%. The models perform comparably, but the low
Specificity values (62.84% to 70.93%) suggest difficulty in correctly
classifying negative instances. Applying 5-minute median filtering
improves the performance metrics. The XGBoost model tops the charts with
an F1 Score of 79.22% and NPV of 74.00%. However, Specificity still
remains low, with values between 54.68% and 74.84% across all models.
With 10-minute median filtering, the metrics improve further. The
XGBoost model still leads with an F1 Score of 80.87% and an NPV of
75.76%. But, Specificity remains a concern, ranging from 57.47% to
76.35% across all models.

@@

The analysis of precision-recall and ROC curves across different models
and ZM prediction types shows varying performance. In terms of
precision-recall AUC, the Decision Tree model consistently outperforms
others, indicating its superior predictive accuracy (see
@fig-pr_curves). Conversely, the Neural Network model generally shows
weaker performance. However, for ROC AUC, the XGBoost model consistently
excels across all data types, indicating a strong ability to
differentiate between classes, while the Neural Network model tends to
underperform (see @fig-roc_curves). The F-measure (F1 score) shows
variable performance across different configurations but generally, the
Decision Tree model yields higher scores.

![Precision-recall curves of the models evaluated across the different
ZM predictions, including raw ZM predictions, as well as 5-minute and
10-minute median smoothing of the ZM raw predictions. The x-axis of the
plot represents the proportion of true wake epochs that were correctly
classified as wake, while the y-axis represents the proportion of all
epochs labeled as wake by the classifier that were classified correctly.
The area under the curve values are displayed as color-coded text in the
plot indicate the area under the Precision-Recall curve for each model
and condition.](visuals/plot_sleep_pr.pdf){#fig-pr_curves}

![Receiver operating characteristic curves of the models evaluated
across the different ZM predictions, including raw ZM predictions, as
well as 5-minute and 10-minute median smoothing of the ZM raw
predictions. The x-axis of the plot represents the proportion of true
asleep epochs that were incorrectly classified as awake, while the
y-axis represents the proportion of all epochs labeled as awake by the
classifier that were correctly classified. The area under the curve
values displayed are displayed as color-coded text in the plot to
indicate the area under the receiver operating characteristic curve for
each model and condition.](visuals/plot_sleep_roc.pdf){#fig-roc_curves}

@tbl-biLSTM_performance presents the performance of the three-class
biLSTM multiclassifier on raw and median-filtered (5-minute and
10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
from 71.36% to 76.04%, indicating overall good performance. Applying
5-minute median filtering improves the metrics further, resulting in F1
Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
and sensitivity. However, Specificity values are not available for this
configuration. With 10-minute median filtering, F1 Scores range from
73.45% to 73.93%, maintaining good performance but with limited
information on Specificity. Further analysis is required to assess
performance across all classes.

@@

## Evaluation of Sleep Quality Summaries

BA plots

pearson regression

scatterplots with identity line and best linear fit

>>>>

---
title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality parameters, including sleep period
time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO).

@@

The current study leverages data from the SCREENS
project[@rasmussen2020], a study conducted from October 2018 to March
2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we focused on
data from child participants aged between 6 and 10 years within the
SCREENS cohort. Our primary sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
utilized for accurate sleep detection, showcasing its suitability for
in-home monitoring[@kaplan2014], while the Z-PLUS effectively
differentiates sleep stages, as evidenced by its alignment with expert
evaluations using PSG data[@wang2015]. In the current study, we treated
all sleep stages as a single category effectively deducing the output of
the ZM to "awake" and "asleep" as the ability to distinguish sleep
stages are not a necessity to derive sleep quality parameters of
interest and to simplify the learning process of the models.

@fig-flow illustrates the selection criteria applied to the children's
recordings from the SCREENS study. We included only ZM recordings that
were accompanied by complete accelerometer data and lasted between 7 and
14 hours. Any night when the ZM reported sensor issues was excluded. The
children whose recordings were considered had an average age of 9.4
years, with a standard deviation of 2.1. In their raw form, the ZM
predictions encompassed 696,779 epochs, each 30 seconds long. Notably,
approximately 84% of the total ZM recording duration was classified as
sleep, resulting in an imbalance of the dataset.

![Flowchart of eligible ZM recording nights included in the
study](visuals/flowchart_of_elligible_nights.pdf){#fig-flow}

Finally, we affirm that the SCREENS study received approval from the
Regional Scientific Committee of Southern Denmark, and all data handling
processes complied with the General Data Protection Regulation (GDPR),
ensuring the ethical and secure management of participant information.

@@

In this study, data processing of the raw accelerometer data began with
a low-pass filtration step using a 4th order Butterworth filter with a 5
Hz cut-off frequency to eliminate high-frequency noise. Following
filtration, data were partitioned into overlapping 2-second intervals,
each successive interval sharing a 50% overlap with the previous one
similar to methods described by Skotte et al.[@skotte_detection_2014].
Any non-wear data was remove using previously described
methods[@skovgaard2023] and data was resampled to 30-second epochs so
every sample classified by the algorithms corresponds to a 30-second
epoch scored during the ZM recordings. Subsequently, we performed a
feature extraction process that yielded a set of 88 features, providing
a robust characterization of the data. Extracted from accelerometer and
temperature signals, these features include temporal elements that use
both lag and lead values, capturing dynamic data trends by incorporating
measurements from preceding and upcoming intervals. Furthermore,
inspired by Walch et al.[@walch2019], we incorporated sensor-independent
features to encapsulate circadian rhythms. These features offer unique
insights not directly discernible from sensor outputs and are meant to
approximate the changing drive of the circadian clock to sleep over the
course of the night (see @fig-sensor-independent). Furthermore, the
feature set was enriched by including signal characteristics, which
encompass vector magnitude, mean crossing rate, skewness, and kurtosis
for each of the x, y, and z dimensions. All features are summarized in
table ??? **in the supplementary matrials**. Subsequently, we merged the
ZM and corresponding accelerometer recordings. Any overlapping time
between the ZM and accelerometer data was treated as 'in-bed' time, with
the remaining time considered 'out-of-bed'. This process yielded a
dataset providing a around the clock temporal view of each participant's
activity and sleep patterns.

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings most likely occurring at the end of each
cycle[@galland_normal_2012]. Upon examining the raw ZM predictions, we
noted a significant overestimation in the number of awakenings per night
for the children in our study, exceeding what would be expected based on
typical sleep cycle patterns. In particular, many of these brief
awakenings could be considered as noise, which when present in the data,
can potentially hinder the learning process of machine learning models
by obscuring the underlying patterns that the models are trying to
learn, leading to less accurate predictions. Consequently, we elected to
train and evaluate our models using not only the raw ZM output, but also
versions that were subjected to 5-minute and 10-minute median filters.
This approach, by mitigating this noise, resulted in an anticipated,
more age-appropriate count of awakenings per night, providing a more
accurate depiction of children's sleep patterns (see @fig-zm-median).

@@

In this study, we employed two distinct modeling strategies to analyze
sleep patterns from thigh-mounted accelerometer data. We used a
sequential strategy, comprising an ensemble of four pairs of models,
each pair featuring the same algorithm. This strategy aimed to make the
prediction task more manageable for the algorithms by breaking it down
into a sequence of two binary classifications: first predicting 'in-bed'
time, then 'sleep' time. Simultaneously, we also used a multiclass
approach utilizing a bidirectional Long Short-Term Memory
(biLSTM)[@hochreiter1997] neural network.

@@

In this study, we employed a biLSTM, a multiclass classifier, to predict
three distinct states: out-of-bed-awake, in-bed-awake, and
in-bed-asleep. The architecture of the biLSTM was set up with four
layers, each equipped with 128 hidden units. This configuration was
intentionally chosen to balance between model complexity and training
efficiency: it provided the depth necessary for learning intricate
patterns while remaining feasible for timely training. The bidirectional
design of the LSTM served to enhance data interpretation and mitigate
overfitting by doubling the hidden units at each time step. For input,
we used tensors shaped as sequences, with each sequence spanning 10
minutes and a step size of one.This approach follows in the footsteps of
previous studies that utilized LSTM models for sleep detection. These
studies showcased the promising potential of LSTMs in capturing complex
temporal patterns. Particularly, the works of Sano et al. [@sano2019]
and Chen et al. [@chen2021] demonstrated the effectiveness of LSTM
models in improving sleep detection using accelerometer data,
underscoring the value of this modeling approach.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. This division ensured that samples from the same night
were never simultaneously present in both sets. To optimize
hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
a regular grid, comprising 20 different combinations of hyperparameters.
The F1 score served as the optimization metric. The best-performing set
of hyperparameters was then used to fit the models to the full training
dataset. This approach allowed us to maximize performance by leveraging
all available data. Moreover, after extracting the in-bed time from the
initial sequential models, the imbalance on the resulting dataset could
cause biases during model training, as models may favor predicting the
majority class. To rectify this, we employed the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
samples by interpolating random samples with their nearest neighbors. We
utilized the themis R package[@themis] to implement SMOTE, resulting in
a balanced distribution of training samples across both classes.

The biLSTM model was trained to differentiate between three states:
out-of-bed-awake, in-bed-awake, and in-bed-asleep. The data used for
training the biLSTM was randomly divided into training, validation, and
test sets, based on a 50/25/25 split. We ensured that data from the same
night was not present across different sets. The model was trained using
the Adam optimizer, selected for its computational efficiency and
adaptability of the learning rate during training. Given the multiclass
classification task with mutually exclusive classes, we employed the
cross-entropy loss function. To obtain a probability distribution over
the classes, the softmax activation function was applied at the output
layer. We evaluated the model's performance using the F1 score on both
the training and validation sets. We implemented early stopping with a
patience of 3 epochs, halting the training process if there was no
improvement in the validation loss over three consecutive epochs.

### Model Validation

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs. out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs. awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. To further
illustrate model performance, we provide confusion matrices for the full
dataset, encompassing both in-bed and out-of-bed data. These matrices
report relative counts, column percentages (the proportion of the true
class accurately predicted), and row percentages (the proportion of
predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research[@hjorth2012; @kushida2001].

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the limits of agreement as the mean
difference plus or minus 1.96 times the standard deviation of these
differences. To ensure our measurements were robust and accounted for
intra-subject variability, we estimated the 95% confidence intervals for
both the bias and the limits of agreement using a bias-corrected and
accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
sleep quality parameteres included the following:

@@

@@

As reported in @tbl-zm_overview the sleep quality parameters derived
from ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. SPT were consistent across raw and filtered
datasets (mean: 9.2 ± 2.1 hours), corresponding to the length of the ZM
recording. TST and SE increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. LPS
also elevated, suggesting the filter smooths out brief awakenings at
sleep onset, leading to a prolonged time to persistent sleep. A
significant change was seen in WASO, which dropped from 39.0 ± 33.6
minutes in raw data to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in
the 5-minute and 10-minute filtered data, respectively. The number of
awakenings was also considerably reduced with the application of
filters. In the raw data, the average number of awakenings was 34.46 ±
11.33 per night, which reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the
5-minute and 10-minute filtered data sets respectively.

@@

## Performance on Epoch-to-Epoch Basis

The epoch-to-epoch evaluation of predicting in-bed time is outlined in
@tbl-in_bed_performance, and demonstrates practically equivalent
performance across all model types. The F1 score ranges from 94.4%
(Decision Tree) to 95.4% (XGBoost), while accuracy ranges from 95.3%
(Decision Tree) to 96.1% (XGBoost). Sensitivity, Precision, and
Specificity also demonstrate consistent results across the models. The
XGBoost model, despite recording the highest metrics with an F1 score of
95.4% and accuracy of 96.1%, outpaced the others only marginally.

@@

@tbl-sleep_performance details the performance of the second sequential
models on raw and median-filtered (5 and 10 minute) ZM predictions.
These are the predictions of the sequential models' ability to predict
asleep/awake on only the extracted in-bed time. The F1 scores, which are
unweighted macro averages, for raw ZM predictions range from 71.05% to
76.18%. The models perform comparably, but the low specificity values
(62.84% to 70.93%) suggest difficulty in correctly classifying awake
epochs. Applying 5-minute median filtering improves the performance
metrics. The XGBoost model tops the charts with an F1 score of 79.22%
and NPV of 74.00%. However, specificity still remains low, with values
between 54.68% and 74.84% across all models. With 10-minute median
filtering, the metrics improve further. The XGBoost model still leads
with an F1 score of 80.87% and an NPV of 75.76%. But, specificity
remains low, ranging from 57.47% to 76.35% across all models.

@@

The complete set of confusion matrices generated from data both
containing the out-of-bed and in-bed time are presented in
@fig-conf_mat. The figure shows favorable epoch-to-epoch performance
across across all sequential models, however, it is evident that the
biLSTM is less successful in classifying the in-bed-awake class which
cannot be deduced from the confusion matrices from the sequential
models. **\[fejl i biLSTM confmat. Alle mats er ens\]**

![Confusion matrices for binary sleep prediction. The middle of each
tile is the normalized count (overall percentage) and, beneath it, the
count. The bottom number is the column percentage (target). At the right
side of each tile is the row percentage (prediction). i) decision tree,
ii) logistic regression, iii) feed-forward neural net, iv) xgboost, and
v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}

@tbl-biLSTM_performance presents the performance of the three-class
biLSTM multiclassifier on raw and median-filtered (5-minute and
10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
from 71.36% to 76.04%, indicating overall good performance. Applying
5-minute median filtering improves the metrics further, resulting in F1
Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
and sensitivity. **\[Overvej flere metrics. Evt. per class... pt er det
lidt svært at sammenligne på tværs af lstm og andre modeller ud over
F1.\]** With 10-minute median filtering, F1 Scores range from 73.45% to
73.93%, maintaining good performance but with limited information on
Specificity. Further analysis is required to assess performance across
all classes.

@@

## Evaluation of Sleep Quality Parameters

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The eed-forward neural network also exhibited
similar bias as the decision tree and the logistic regression models,
but with a higher overestimation in WASO. On the other hand, the XGboost
model showed least bias among all, especially in its 5-minute median
predictions. Considering LOA, the decision tree had higher variability
across different sleep quality parameters and filtering techniques,
particularly for LPS and WASO, which indicates lower agreement with ZM.
Other models had comparable LOA but with notable exceptions. For
example, TST LOA for the logistic regression model was particularly wide
in the 5-minute median predictions. Correlation-wise, the pearson
coefficient, revealed that the XGboost model consistently had the
highest correlation with ZM across all sleep qualityparameters and
filtering methods Notably, the XGboost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

```{r}
#| echo: false
#| message: false
#| label: tbl-ba_cor
#| tbl-cap: Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors. Each value is provided with its 95% Confidence Interval (CI).

source("code/table_ba_cor_statistics.R")

short_tbl_ba_cor

```
